Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-24 bbbq_article.tex
Use TMP for 'transmembrane protein'

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions. 
We obtained 1,129 gene names associated with the phrase 'membrane protein'.
These genes are linked to 4,811 protein isoforms, 
of which 2,568 are predicted to be transmembrane proteins.
We obtained 38,882 SNPs that resulted in an
amino acids substitution, of which 21,738 were located 
in predicted transmembrane proteins.
Per protein, we calculated two percentages: 
(1) the percentage of the gene coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without membrane-associated proteins,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

To statistically test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution.
We calculated the number of SNPs that are expected to be
found in TMHs by chance, $E(n_{SNP\_TMH})$, when including
and excluding soluble proteins (with symbols
$E(n_{SNP\_TMH}^{+})$ and $E(n_{SNP\_TMH}^{-})$ respectively).
The number of SNPs that are expected to be
found in TMHs by chance when including soluble proteins,
$E(n_{SNP\_TMH}^{+})$ 
equals the number of (all) SNPs, $n_{SNP}^{+} = 38,882$, multiplied by the fraction
of all proteins that are TMH, $f_{TMH}^{+} = 9.96\%$,
resulting in 
$E(n_{SNP\_TMH}^{+}) = n_{SNP}^{+} \cdot f_{TMH}^{+} \approx 3873$
The number of SNPs that are expected to be
found in TMHs by chance when excluding soluble proteins,
$E(n_{SNP\_TMH}^{-})$ 
equals the number of SNPs in TMHs, $n_{SNP}^{-} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH}^{-} = 18.66\%$,
$E(n_{SNP\_TMH}^{-}) = n_{SNP}^{-} \cdot f_{TMH}^{-} \approx 4056$.
The discrepancy between the two expectations
is caused by the distribution of SNPs over the soluble and
transmembrane proteins: 
there are 38,882 SNPs found in 4811 proteins (i.e. $\approx 8.08$ SNP per protein),
of which 21,738 SNPs in 2568 integral membrane proteins (i.e. $\approx 8.46$
SNP per integral membrane protein).

@@

In this experiment we predicted epitopes that overlap with 
TMHs from a human, bacterial and viral proteome,
would these proteins be expressed in a human host.
Bacteria, however have different cell membranes and cell walls, 
hence different structural requirements for a TMH.
Both topology prediction tools were trained to recognize
human TMHs, thus we cannot be sure that
the transmembrane regions predicted in bacterial proteins
are actually part of an integral membrane protein.
For the purpose of this study, we assume the 
error in topology predictions to be unbiased way towards topology.
In other words: that a bacterial TMH is incorrectly
predicted to be absent just as often as it is incorrectly
predicted to be present elsewhere.

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_n_tmhs_per_protein.png}
  \caption{
    Histogram of the number of TMHs predicted per protein,
    for the integral membrane proteins used.
  }
  \label{fig:n_tmhs_per_protein}
\end{figure}

>>>>

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions. 
We obtained 1,129 gene names associated with the phrase 'membrane protein'.
These genes are linked to 4,811 protein isoforms, 
of which 2,568 are predicted to be transmembrane proteins (TMPs),
where the others are predicted to be membrane-associated proteins.
We obtained 38,882 SNPs that resulted in an
amino acids substitution, of which 21,738 were located 
in predicted transmembrane proteins.
Per protein, we calculated two percentages: 
(1) the percentage of the gene coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without membrane-associated proteins,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

To statistically test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution.
We calculated the number of SNPs that are expected to be
found in TMHs by chance, $E(n_{SNP\_TMH})$, when including
and excluding soluble proteins (with symbols
$E(n_{SNP\_TMH}^{+})$ and $E(n_{SNP\_TMH}^{-})$ respectively).
The number of SNPs that are expected to be
found in TMHs by chance when including soluble proteins,
$E(n_{SNP\_TMH}^{+})$ 
equals the number of (all) SNPs, $n_{SNP}^{+} = 38,882$, multiplied by the fraction
of all proteins that are TMH, $f_{TMH}^{+} = 9.96\%$,
resulting in 
$E(n_{SNP\_TMH}^{+}) = n_{SNP}^{+} \cdot f_{TMH}^{+} \approx 3873$
The number of SNPs that are expected to be
found in TMHs by chance when excluding soluble proteins,
$E(n_{SNP\_TMH}^{-})$ 
equals the number of SNPs in TMHs, $n_{SNP}^{-} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH}^{-} = 18.66\%$,
$E(n_{SNP\_TMH}^{-}) = n_{SNP}^{-} \cdot f_{TMH}^{-} \approx 4056$.
The discrepancy between the two expectations
is caused by the distribution of SNPs over the soluble and
transmembrane proteins: 
there are 38,882 SNPs found in 4811 proteins (i.e. $\approx 8.08$ SNP per protein),
of which 21,738 SNPs in 2568 TMPs (i.e. $\approx 8.46$
SNP per TMP).

@@

In this experiment we predicted epitopes that overlap with 
TMHs from a human, bacterial and viral proteome,
would these proteins be expressed in a human host.
Bacteria, however have different cell membranes and cell walls, 
hence different structural requirements for a TMH.
Both topology prediction tools were trained to recognize
human TMHs, thus we cannot be sure that
the transmembrane regions predicted in bacterial proteins
are actually part of a TMH.
For the purpose of this study, we assume the 
error in topology predictions to be unbiased way towards topology.
In other words: that a bacterial TMH is incorrectly
predicted to be absent just as often as it is incorrectly
predicted to be present elsewhere.

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_n_tmhs_per_protein.png}
  \caption{
    Histogram of the number of TMHs predicted per protein,
    for the trans-membrane proteins used.
  }
  \label{fig:n_tmhs_per_protein}
\end{figure}
